# Non-Communicable Disease Burden (NCD)

```{r echo=FALSE, message=FALSE}
# Load necessary packages
pacman::p_load(tidyverse, knitr, here, janitor)

# Import the non-communicable diseases data
data_ncd <- read_csv(here("data", "burden-of-disease-ncd.csv"))
```

## Table of Estimates for NCD Burden Over Time

```{r echo=FALSE, message=FALSE}
# Here render a table for the DALY burden over time for the three countries 
# You should pivot the data to show each country in a separate column.
# Use kable() from the knitr package to render the table.

data_ncd2 <- data_ncd %>%
  rename(DALYS = `DALYs (Disability-Adjusted Life Years) - Non-communicable diseases - Sex: Both - Age: Age-standardized (Rate)`) %>% 
  filter(Entity %in% c("Nigeria", "Jamaica", "Ghana"))

 #data_ncd2
```

```{r echo=FALSE}
ncd_data <- data_ncd2 %>% 
  mutate(DALYS = round(DALYS)) %>% 
  select(Entity, Year, DALYS) %>% 
  pivot_wider(names_from = Entity, values_from = DALYS)
 kable(ncd_data, align = "lccc", "pipe", caption ="Table of Estimates for NCD Burden Over Time" )
```

## Plot Showing Trends in NCD Burden Over Time

```{r echo=FALSE}

ggplot(data_ncd2) +
  aes(x = Year, y = DALYS, colour = Entity) +
  geom_point(shape = "circle", size =2) +
  scale_color_hue(direction = 1) +
  theme_minimal() +
  labs(title = "Trends in NCD Burden Over Time",
       caption = "Figure 1",
       x = "Years",
       y = "Disability-Adjusted Life Years (DALY)",
       color = "Country")
```

## Summary of NCD Burden Findings

In the 30-year period shown in Figure 1 above, the burden of Non-Communicable Diseases (NCDs) continues to drop at a very low rate for Nigeria and Ghana. While Jamaica may not have comparative levels of NCDs with Nigeria and Ghana, a precipitous drop in the 5-year period between 2000 and 2005 only led to a similar rate of increase over the next five years to 2010, where it has since plateaued. Given that the data is looking at NCDs, it may be difficult to adduce a reason for the increase.

### Cumulative DALY Burden for NCDs

```{r echo=FALSE}
cum_daly <- data_ncd2 %>%
  group_by(Entity) %>% 
  mutate(cumul_daly = round(cumsum(DALYS)))
#cum_daly
```

```{r echo=FALSE}
ggplot(cum_daly) +
  aes(x = Year, y = cumul_daly, colour = Entity) +
  geom_point(shape = "circle", size =2) +
  scale_y_continuous(labels = scales::comma)+
  scale_color_hue(direction = 1) +
  theme_minimal() +
  labs(title = "Cumulative Trends in NCD Burden Over Time",
       caption = "Figure 2",
       x = "Years",
       y = "Cumul. DALY",
       color = "Country")
```

The cumulative trends show that the burden of NCDs in the three countries has grown exponentially over the 30 years covered in the analysis. The trends in Ghana and Nigeria are similar, with just a slight difference from Jamaica.
